Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-12-21 manuscript/manuscript.Rmd
references updates.

This paper presents the Perception-Process-Product (hereafter 'Triple P') conceptual framework to expand the scope of experimental archaeology. The field has long tended to adopt the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, and center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. When applied to the experimental study of ancient craftsmanship, it implies that a technological solution that is simpler to reproduce is more likely to be the one used in the archaeological context. Nevertheless, Occam's razor, or the law of parsimony, can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use. The two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* (i.e., the idea that humans are consistently rational and narrowly self-interested agents pursuing for optimality) for many anthropologists [@henrichSearchHomoEconomicus2001]. Parallelly, Heyes [@heyes2012] also questioned the abuse of parsimony in animal behavioural research and further proposed that new observational and experimental studies that allow differential predictions to be tested become necessary when both a simple and a complex mechanism can explain the phenomenon of interest.

@@

Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artefacts. Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions. Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@montgomery2023; @marshall2002; @douglass2020].

@@

Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019]. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artefacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage. Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely 'an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen).' In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography become necessary.

>>>>

This paper presents the Perception-Process-Product (hereafter 'Triple P') conceptual framework to expand the scope of experimental archaeology. The field has long tended to adopt the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009; @schmidt2019; @blessing2021], whether explicitly or implicitly, and center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. When applied to the experimental study of ancient craftsmanship, it implies that a technological solution that is simpler to reproduce is more likely to be the one used in the archaeological context. Nevertheless, Occam's razor, or the law of parsimony, can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use. The two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* (i.e., the idea that humans are consistently rational and narrowly self-interested agents pursuing for optimality) for many anthropologists [@henrichSearchHomoEconomicus2001]. Parallelly, Heyes [@heyes2012] also questioned the abuse of parsimony in animal behavioural research and further proposed that new observational and experimental studies that allow differential predictions to be tested become necessary when both a simple and a complex mechanism can explain the phenomenon of interest.

@@

Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artefacts. Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022; @milks2023] across different regions. Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@montgomery2023; @marshall2002; @douglass2020].

@@

Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as cognitive affordances [@hussain2021; @roepstorff2008], sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019]. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artefacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage. Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely 'an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen).' In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography become necessary.
